Efficiently stable presentations from error-correcting codes

### 2023-11-01 efficient_stability.tex
added a few comments, made small edits in intro

\mnote{Here I genuinely changed the phrasing. I don't think it is currently very good, but it is better than it was. Feel free to iterate:}\tnote{slightly edited}With this goal in mind we analyze various natural examples of presentations. We provide a general method for constructing presentations of $\Z_2^k$ from linear error-correcting codes. We observe that the resulting presentation has a weak form of stability exactly when the code is  \emph{locally testable}.\mnote{I have a problem with the notion of locally testable. The notion reffers ususally to testable LDPCs. We do not need the low density to define our notions and construction, only the testable part. I think we should change it to just testable codes... What do you think? This appears throughout the intro... It won't be a major change}\tnote{Right, it's not clear what motivation there is for the ``locally'' aspect in the language of groups; that the relations be ``short''---is that ever relevant? I'm ok for dropping ``locally''. It's not hard as we can search for the word.} This raises the question of whether locally testable codes give rise to genuinely stable presentations using this method.  While we cannot show that this is the case in general, we leverage recent results in the study of non-local games in quantum information theory (Ji et al., Discrete Analysis 2021) to show that a specific instantiation of our construction, based on the Reed-Muller family of codes, leads to a stable presentation of $\Z_2^k$ of size $\poly\log(k)$ only. As an application, we combine this result with recent work of de la Salle (arXiv:2204.07084) to re-derive  the quantum low-degree test of Natarajan and Vidick (IEEE FOCS'18), which is a key building block in the recent refutation of Connes' Embedding Problem via complexity theory (Ji et al., arXiv:2001.04383). 
\end{abstract}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? 
One can verify that with $\mu_R$ which chooses with probability $\nicefrac{1}{3}$ whether to check an involution, a commutation or a matrix row,  and $\mu_S$ the uniform distribution, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=3\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer cannot be straightforward: even deducing basic properties about the group $G(h)$ given its presentation -- let alone its stability properties -- appears to be a challenging task. For example, there are examples of parity check matrices $h$ for which the group $G(h)$ is non-abelian or even non-amenable; see Remark~\ref{rk:non-abelian}. Actually, every finitely generated group can be embedded into $G(h)$ for some $h$ (see \cite{slofstra2019set}).

@@

We do not know of any other presentation of $\Z_2^k$, arguably the simplest example one could think of, that is stable with similar parameters as the ones stated in the theorem. It would be very interesting to discover different such presentations, built from locally testable error-correcting codes or not, for this group or others.

@@

Now we see that to each matrix $h$ we have associated a group $G(h)$, and a game $\game_h$. Moreover, and quite interestingly, there is a one-to-one correspondence between representations of $G(h)$ and \emph{perfect strategies} in $\game_h$, i.e.\ strategies that have success probability $1$ in the game.\footnote{This correspondence was established for finite-dimensional strategies, and finite-dimensional representations, in~\cite{cleve2014characterization}. Extensions to infinite-dimensional strategies and representations have appeared in~\cite{cleve2017perfect}. See also e.g.~\cite{kim2018synchronous} for generalizations to the broader class of \emph{synchronous games}.}
This correspondence enables one to ``embed'' group representations into quantum strategies, thereby forcing them to demonstrate a high level of complexity; this is the approach at the heart of~\cite{slofstra2019set}. Going further, for applications one is often required to understand not only optimal but also near-optimal strategies, whose success probability is e.g.\ $1-\eps$. The same correspondence associates to such strategies approximate homomorphisms of $G(h)$~\cite{slofstra2018entanglement}. Here once can see that measuring closeness on average over the choice of relation is a natural choice, which is all but forced by the definition of a game, where the questions are selected according to some pre-specified distribution.

@@

\paragraph{Open questions.}\tnote{added this paragraph}
We leave many questions open. A natural direction is to determine if there are simple sufficient conditions on the matrix $h$ that guarantee that the presentation $G(h)$ is stable. As we discussed, the condition that $\ker h$ is a testable code is equivalent to stability for $1$-dimensional permutations. It thus seems likely that testability is not a sufficient condition in general; can testability be determined by a combinatorial parameter of $h$? More generally, finding other examples of efficient stability seems of intrinsic interest, besides potential applications to property testing and the construction of nonlocal games. In a different direction, as mentioned in the previous paragraph an analogue of Theorem~\ref{thm:main-inf} for the case of representations in permutations could have important implications towards showing the existence of non-sofic groups~\cite{CL_part1,CL_part2,BCLV_subgroup_tests}\tnote{Michael, can you add a reference to the relevant paper of yours/ours?}\mnote{Done. But, it seems as if the second paragraph of the 'related works on stability' subsection already hints towards this open problem.}\tnote{Ok thanks, I reformulated. We could also remove the refs here and only mention them in the previous paragraph. I think it's ok but if you prefer that way you can edit.}

>>>>

\mnote{Here I genuinely changed the phrasing. I don't think it is currently very good, but it is better than it was. Feel free to iterate:}\tnote{slightly edited}With this goal in mind we analyze various natural examples of presentations. We provide a general method for constructing presentations of $\Z_2^k$ from linear error-correcting codes. We observe that the resulting presentation has a weak form of stability exactly when the code is  \emph{locally testable}.\mnote{I have a problem with the notion of locally testable. The notion reffers ususally to testable LDPCs. We do not need the low density to define our notions and construction, only the testable part. I think we should change it to just testable codes... What do you think? This appears throughout the intro... It won't be a major change}\tnote{Right, it's not clear what motivation there is for the ``locally'' aspect in the language of groups; that the relations be ``short''---is that ever relevant? I'm ok for dropping ``locally''. It's not hard as we can search for the word.} \hnote{I'm fine with dropping ``locally''. The key thing that makes the problem nontrivial is that the \emph{number} of relations (i.e. complexity of presentation) is small, right?} This raises the question of whether locally testable codes give rise to genuinely stable presentations using this method.  While we cannot show that this is the case in general, we leverage recent results in the study of non-local games in quantum information theory (Ji et al., Discrete Analysis 2021) to show that a specific instantiation of our construction, based on the Reed-Muller family of codes, leads to a stable presentation of $\Z_2^k$ of size $\poly\log(k)$ only. As an application, we combine this result with recent work of de la Salle (arXiv:2204.07084) to re-derive  the quantum low-degree test of Natarajan and Vidick (IEEE FOCS'18), which is a key building block in the recent refutation of Connes' Embedding Problem via complexity theory (Ji et al., arXiv:2001.04383). 
\end{abstract}

@@

We can now ask the question: is $G(h)$, the group presentation defined in~\eqref{eq:gh-intro}, stable according to Definition~\ref{def:eff-stab-intro}? 
One can verify that with $\mu_R$ which chooses with probability $\nicefrac{1}{3}$ whether to check an involution, a commutation or a matrix row,  and $\mu_S$ the uniform distribution, $G(h)$ is $(\delta,\mu_S,\mu_R,1)$-stable for $\delta=3\eps/\rho$ if and only if $\code = \ker h$ is locally testable with soundness $\rho>0$. But what about higher-dimensional approximate representations? Are these also stable, or does one need to make further requirements on $\code$ beyond local testability? We do not yet have a comprehensive answer to these questions. However, the answer cannot be straightforward: even deducing basic properties about the group $G(h)$ given its presentation -- let alone its stability properties -- appears to be a challenging task. For example, there are examples of parity check matrices $h$ for which the group $G(h)$ is non-abelian or even non-amenable; see Remark~\ref{rk:non-abelian}. In fact, every finitely generated group can be embedded into $G(h)$ for some $h$ (see \cite{slofstra2019set}).

@@

We do not know of any other presentation of $\Z_2^k$, arguably one of the simplest groups one could think of, that is stable with similar parameters as the ones stated in the theorem. It would be very interesting to discover different such presentations, built from locally testable error-correcting codes or not, for this group or others.

@@

Now we see that to each matrix $h$ we have associated a group $G(h)$ \hnote{do we have to use the tilde version now?}, and a game $\game_h$. Moreover, and quite interestingly, there is a one-to-one correspondence between representations of $G(h)$ and \emph{perfect strategies} in $\game_h$, i.e.\ strategies that have success probability $1$ in the game.\footnote{This correspondence was established for finite-dimensional strategies, and finite-dimensional representations, in~\cite{cleve2014characterization}. Extensions to infinite-dimensional strategies and representations have appeared in~\cite{cleve2017perfect}. See also e.g.~\cite{kim2018synchronous} for generalizations to the broader class of \emph{synchronous games}.}
This correspondence enables one to ``embed'' group representations into quantum strategies, thereby forcing them to demonstrate a high level of complexity; this is the approach at the heart of~\cite{slofstra2019set}. Going further, for applications one is often required to understand not only optimal but also near-optimal strategies, whose success probability is e.g.\ $1-\eps$. The same correspondence associates to such strategies approximate homomorphisms of $G(h)$~\cite{slofstra2018entanglement}. Here once can see that measuring closeness on average over the choice of relation is a natural choice, which is all but forced by the definition of a game, where the questions are selected according to some pre-specified distribution.

@@

\paragraph{Open questions.}\tnote{added this paragraph}
We leave many questions open. A natural direction is to determine if there are simple sufficient conditions on the matrix $h$ that guarantee that the presentation $G(h)$ (\hnote{added:} or $\widetilde{G(h)}$) is stable. As we discussed, the condition that $\ker h$ is a testable code is equivalent to stability for $1$-dimensional permutations. It thus seems likely that testability is not a sufficient condition in general; can testability be determined by a combinatorial parameter of $h$? More generally, finding other examples of efficient stability seems of intrinsic interest, besides potential applications to property testing and the construction of nonlocal games. In a different direction, as mentioned in the previous paragraph an analogue of Theorem~\ref{thm:main-inf} for the case of representations in permutations could have important implications towards showing the existence of non-sofic groups~\cite{CL_part1,CL_part2,BCLV_subgroup_tests}\tnote{Michael, can you add a reference to the relevant paper of yours/ours?}\mnote{Done. But, it seems as if the second paragraph of the 'related works on stability' subsection already hints towards this open problem.}\tnote{Ok thanks, I reformulated. We could also remove the refs here and only mention them in the previous paragraph. I think it's ok but if you prefer that way you can edit.}
